|

### 2021-11-12 submission/manuscript.rmd
update abstract

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its interaction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the Ti's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

>>>>

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting
 individualized treatment effects in RCTs. **Study Design and
 Setting:** We simulated data using diverse assumptions for a baseline
 prognostic index of risk (PI), the shape of its interaction with
 treatment (none, linear, quadratic or non-monotonic) and the magnitude
 of treatment-related harms. In each sample we predicted absolute
 benefit using: models with a constant relative treatment effect;
 stratification in quarters of the PI; models including a linear
 interaction of treatment with the PI; models including an interaction
 of treatment with a restricted cubic spline (RCS) transformation of
 the PI; an adaptive approach using Akaike’s Information Criterion. We
 evaluated predictive performance using root mean squared error and
 measures of discrimination and calibration for benefit. Starting from
 a base scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
 the PI 0.75), we varied the sample size, the treatment effect size,
 and the PI's AUC. **Results:** Models including a PI by treatment
 interaction performed best with smaller sample sizes and/or lower AUC
 of the PI. Otherwise, RCS (3 knots) models proved more robust for the
 majority of the deviation settings. The adaptive approach was unstable
 with smaller sample sizes. **Conclusion:** Depending on the setting,
 the linear interaction or the RCS (3 knots) model should be preferred.

authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

The results from all individual scenarios can be explored online at
[https://arekkas.shinyapps.io/simulation_viewer/](https://arekkas.shinyapps.io/simulation_viewer/).
